---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# mealprepR

<!-- badges: start -->

[![R build status](https://github.com/UBC-MDS/mealprepR/workflows/R-CMD-check/badge.svg)](https://github.com/UBC-MDS/mealprepR/actions) 

[![codecov](https://codecov.io/gh/UBC-MDS/mealprepR/branch/master/graph/badge.svg)](https://codecov.io/gh/UBC-MDS/mealprepR)


<!-- badges: end -->

MealprepR offers a toolkit, made with care, to help users save time in the data preprocessing kitchen.

## Overview

Recognizing that the preparation step of a data science project often requires the most time and effort, `mealprepR` aims to help data science chefs of all specialties master their recipes of analysis. This package tackles pesky tasks such as classifying columns as categorical or numeric ingredients, straining NA values and outliers, and automating a preprocessing recipe pipeline.

## Functions

`find_fruits_veg()`: This function returns the indices of numeric and categorical variables in the dataset.

`find_missing_ingredients()`: For each column with missing values, this function will create a reference list of row indices, sum the number, and calculate the proportion of missing values.

`find_bad_apples()`: This function uses a univariate approach to outlier detection. For each column with outliers (values that are 2 or more standard deviations from the mean), this function will create a reference list of row indices with outliers, and the total number of outliers in that column.

`make_recipe()`: This function is used to quickly apply common data preprocessing techniques.

## mealprepR and R's Ecosystem

**mealprepR** complements many of the existing packages in the R ecosystem around the theme of data preprocessing. With respect to categorizing the columns of a dataframe by type, the base R code `sapply(yourdataframe, class)` will produce the data types of each column. However, there appears to be no easy way to divide the columns into categorical and numerical groups. `find_fruits_veg()` aims to fill this void.  

In terms of missing values, the package [na.tools](https://cran.r-project.org/web/packages/na.tools/na.tools.pdf) provides a suite of tools to check the number and proportion of missing values in a dataset as well as methods controlling how they may be replaced. The [forecast](https://cloud.r-project.org/web/packages/forecast/forecast.pdf) package includes the function `na.interp()` which provides a similar replacement tool for time series data. The gap between these packages is that neither provides a summary of the missing values including the list of indices where they occur. `find_missing_ingredients()` augments these tools by providing a summary dataframe detailing which columns have missing values, as well as their count and proportion.

The base R `summary()` function is a popular first function to run during the data exploration stage of a project because it returns the mean, median, minimum, and maximum of each variable in a dataset, which allows users to easily see whether there are possible outliers. However, the output of this function does not tell you which rows of data these outliers are found in, or how many outliers are present in the dataframe. Packages like [outliers](https://cran.r-project.org/web/packages/outliers/outliers.pdf) and [OutlierDetection](https://cran.r-project.org/web/packages/OutlierDetection/OutlierDetection.pdf) provide many more ways of detecting outliers using both univariate and multivariate outlier detection methods, and some functions such as the `outlierTest()` function from the [car](https://cran.r-project.org/web/packages/car/car.pdf) package also output the row within a dataframe which has the most extreme observation given a model. The `find_bad_apples()` function provides more detail than the base R `summary()` function for outlier detection, but does not provide multivariate outlier detection like other functions from packages specifically made to detect outliers.

Lastly, there are many great tools in the data science ecosystem for pre-processing data such as [caret](https://cran.r-project.org/web/packages/caret/caret.pdf) in R. However, you may find yourself frequently writing the same lengthy code for common preprocessing tasks (e.g scale numeric features and one hot encode categorical features). `preprocess_recipe()` provides a _shortcut function_ to apply your favourite recipes quickly to preprocess data in one line of code.


## Installation

You can install the released version of mealprepR from
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("mealprepR")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("UBC-MDS/mealprepR")
```

## Examples

### `find_fruits_veg()`

### `find_missing_ingredients()`

### `find_bad_apples()`

### `make_recipe()`

Do you find yourslef constantly applying the same data preprocessing techniques time and time again? `make_recipe` can help by applying your favourite preprocessing recips in only a few lines of code.

Below `make_recipe` applies the following common recipe in only one line of code:

1. Split data into training, validation, and testing
2. Standardise and scale numeric features
3. One hot encode categorical features

First load the classic `mtcars` data set.

```{r message=FALSE}
library(dplyr)

X <- dplyr::as_tibble(mtcars) %>%
 mutate(
    carb = as.factor(carb),
    gear = as.factor(gear),
    vs = as.factor(vs),
    am = as.factor(am)
  )
head(X)
```

Use `make_recipe` to quickly apply split your data and apply your favourite preprocessing techniques!

```{r}
library(mealprepR)

mtcars_splits <- make_recipe(
  X = X, 
  y = "gear", 
  recipe = "ohe_and_standard_scaler", 
  splits_to_return = "train_test"
)
head(mtcars_splits$X_train)
```

Calling `make_recipe` with then return a list containing all of your transformed data.

```{r}
str(mtcars_splits)
```




